% Chapter 6 — our first ZK contract from scratch.
% All source code lives in zk-password/ at the root of the repository.

\section{What We Are Building} \label{sec:ch6-overview}

This chapter builds the simplest possible ZK smart contract on Cardano: a UTxO that can only be unlocked by someone who can prove they know a secret password --- without ever revealing the password on-chain.

The password is hashed using \textbf{Poseidon}, a ZK-friendly hash function. The hash is stored publicly in the datum. To spend the UTxO, the spender generates a Groth16 proof that they know a pre-image of that hash. The Aiken validator checks the proof using the BLS12-381 pairing primitives introduced in Plutus V3. If the proof is valid, the funds are released. If not, the transaction fails.

\begin{highlight}
\textbf{What stays private:} the password. \quad \textbf{What goes on-chain:} the Poseidon hash and the proof.
\end{highlight}

The complete source lives in the \texttt{zk-password/} directory at the root of this repository.

\begin{verbatim}
zk-password/
+-- aiken.toml          <- Aiken project manifest
+-- circuit/
|   +-- circuit.circom  <- Circom circuit
+-- lib/
|   +-- zk_password/
|       +-- groth16.ak  <- Groth16 verifier library
+-- validators/
    +-- password.ak     <- on-chain validator
\end{verbatim}

% ─────────────────────────────────────────────────────────────────────────────
\section{The Circuit} \label{sec:ch6-circuit}

The circuit is intentionally minimal. It has one public input (\texttt{hash}) and one private input (\texttt{pwd}), and it asserts a single constraint: the Poseidon hash of the password must equal the public hash.

\begin{verbatim}
pragma circom 2.1.6;

include "../node_modules/circomlib/circuits/poseidon.circom";

template PasswordLock() {
    signal input hash;   // public  -- stored in the datum
    signal input pwd;    // private -- never leaves the prover

    component h = Poseidon(1);
    h.inputs[0] <== pwd;

    hash === h.out;
}

component main {public [hash]} = PasswordLock();
\end{verbatim}

The \texttt{===} operator defines a constraint. If it is violated, Circom will refuse to generate a valid witness, which means no proof can be produced. The only way to satisfy the constraint is to know \texttt{pwd}.

\subsection*{Compile the circuit}

From inside \texttt{zk-password/}, install circomlib and compile:

\begin{verbatim}
npm install -D circomlib

circom --r1cs --wasm --sym --inspect \
       --prime bls12381 circuit/circuit.circom -o circuit/
\end{verbatim}

Inspect the result:

\begin{verbatim}
snarkjs r1cs info circuit/circuit.r1cs
\end{verbatim}

You should see: 1 private input (\texttt{pwd}), 1 public input (\texttt{hash}), and roughly 240 constraints (mostly from the Poseidon internals).

% ─────────────────────────────────────────────────────────────────────────────
\section{The Trusted Setup} \label{sec:ch6-setup}

Follow the same steps as Chapter 5. For local development you can use a minimal 2-contributor ceremony; for production you should run a proper multi-party ceremony or reuse an existing BLS12-381 Powers of Tau file.

\begin{verbatim}
# Phase 1 — Powers of Tau (circuit-independent)
snarkjs powersoftau new bls12381 12 pot12_0000.ptau -v
snarkjs powersoftau contribute pot12_0000.ptau pot12_0001.ptau \
    --name="dev" -e="some entropy"
snarkjs powersoftau beacon pot12_0001.ptau pot12_beacon.ptau \
    00000000000000000000000000000000 10 -n="Dev beacon"
snarkjs powersoftau prepare phase2 pot12_beacon.ptau pot12_final.ptau -v

# Phase 2 — circuit-specific
snarkjs groth16 setup circuit/circuit.r1cs pot12_final.ptau zkey_0000.zkey
snarkjs zkey contribute zkey_0000.zkey zkey_0001.zkey \
    --name="dev" -e="more entropy"
snarkjs zkey beacon zkey_0001.zkey zkey_final.zkey \
    00000000000000000000000000000000 10 -n="Final beacon"

# Export the verification key
snarkjs zkey export verificationkey zkey_final.zkey verification_key.json
\end{verbatim}

\begin{remark}
We use $2^{12} = 4096$ max constraints here --- more than enough for our 240-constraint circuit and fine for development. Increase the power (e.g.\ to 14) for larger circuits.
\end{remark}

% ─────────────────────────────────────────────────────────────────────────────
\section{Generating a Proof} \label{sec:ch6-prove}

Create \texttt{input.json} with your password and its Poseidon hash. Note: the field element values are large integers that must be quoted:

\begin{verbatim}
{
    "hash": "18586133768512220936620950268035737108927924817073536007144749970837830059131",
    "pwd":  "12345"
}
\end{verbatim}

\begin{remark}
In a real application, the user computes \texttt{Poseidon(pwd)} in the browser using a JavaScript library (\texttt{circomlibjs}) and stores only the resulting hash. The raw password is never sent anywhere.
\end{remark}

Generate the proof:

\begin{verbatim}
snarkjs groth16 fullprove \
    input.json \
    circuit/circuit_js/circuit.wasm \
    zkey_final.zkey \
    proof.json public.json

# Verify off-chain (sanity check)
snarkjs groth16 verify verification_key.json public.json proof.json
\end{verbatim}

Output: \texttt{[INFO] snarkJS: OK!}

% ─────────────────────────────────────────────────────────────────────────────
\section{Preparing the Verification Key for Cardano} \label{sec:ch6-compress}

Cardano only accepts \textbf{compressed} BLS12-381 points. The \texttt{verification\_key.json} from snarkjs uses uncompressed points. You must compress all G1 and G2 points before embedding them on-chain.

Use the compression script from Janus Wallet (or any equivalent tool) to produce \texttt{compressed\_vk.json}. The result will have:

\begin{itemize}
  \item \texttt{vk\_alpha1} --- 48-byte hex string (compressed G1)
  \item \texttt{vk\_beta2} --- 96-byte hex string (compressed G2)
  \item \texttt{vk\_gamma2} --- 96-byte hex string (compressed G2)
  \item \texttt{vk\_delta2} --- 96-byte hex string (compressed G2)
  \item \texttt{vk\_ic} --- array of 48-byte hex strings; 2 elements for our 1-public-input circuit
\end{itemize}

These compressed bytes go directly into the \texttt{VerificationKey} fields in the datum.

% ─────────────────────────────────────────────────────────────────────────────
\section{The Groth16 Verifier Library} \label{sec:ch6-groth16}

The file \texttt{lib/zk\_password/groth16.ak} contains a self-contained Groth16 verifier. The core logic mirrors Janus Wallet's \texttt{zk\_verify} function, adapted for our simpler single-input circuit.

\begin{verbatim}
use aiken/builtin.{
  bls12_381_final_verify, bls12_381_miller_loop,
  bls12_381_mul_miller_loop_result,
}
use aiken/crypto/bls12_381/g1
use aiken/crypto/bls12_381/g2
use aiken/crypto/bls12_381/scalar.{Scalar}

pub type VerificationKey {
  alpha: ByteArray, // G1 — 48 bytes
  beta:  ByteArray, // G2 — 96 bytes
  gamma: ByteArray, // G2 — 96 bytes
  delta: ByteArray, // G2 — 96 bytes
  ic:    List<ByteArray>, // G1 points; length = num_public_inputs + 1
}

pub type Proof {
  pi_a: ByteArray, // G1 — 48 bytes
  pi_b: ByteArray, // G2 — 96 bytes
  pi_c: ByteArray, // G1 — 48 bytes
}
\end{verbatim}

The key function is \texttt{groth16\_verify}. It:

\begin{enumerate}
  \item Decompresses all proof and verification-key points.
  \item Computes $\mathbf{vk\_x} = IC_0 + \sum_i (\text{public\_input}_i \cdot IC_{i+1})$.
  \item Evaluates four Miller loops.
  \item Calls \texttt{bls12\_381\_final\_verify} to check $e(\pi_A, \pi_B) = e(\alpha, \beta) \cdot e(\mathbf{vk\_x}, \gamma) \cdot e(\pi_C, \delta)$.
\end{enumerate}

\begin{verbatim}
pub fn groth16_verify(
  vk: VerificationKey,
  proof: Proof,
  public_inputs: List<Scalar>,
) -> Bool {
  let a = g1.decompress(proof.pi_a)
  let b = g2.decompress(proof.pi_b)
  let c = g1.decompress(proof.pi_c)

  let vk_x =
    when vk.ic is {
      [] -> fail @"vk.ic must not be empty"
      [ic0, ..rest] -> derive(rest, public_inputs, g1.decompress(ic0))
    }

  let e_ab        = bls12_381_miller_loop(a, b)
  let e_alpha_beta =
    bls12_381_miller_loop(g1.decompress(vk.alpha), g2.decompress(vk.beta))
  let e_x_gamma   = bls12_381_miller_loop(vk_x, g2.decompress(vk.gamma))
  let e_c_delta   = bls12_381_miller_loop(c, g2.decompress(vk.delta))

  let rhs =
    bls12_381_mul_miller_loop_result(
      bls12_381_mul_miller_loop_result(e_alpha_beta, e_x_gamma),
      e_c_delta,
    )

  bls12_381_final_verify(e_ab, rhs)
}
\end{verbatim}

% ─────────────────────────────────────────────────────────────────────────────
\section{The Aiken Validator} \label{sec:ch6-validator}

The validator in \texttt{validators/password.ak} is deliberately short. All the cryptographic weight is in the Groth16 library; the validator only wires the datum, redeemer, and the single public input together.

\begin{verbatim}
use aiken/crypto/bls12_381/scalar.{Scalar}
use cardano/transaction.{OutputReference, Transaction}
use zk_password/groth16.{Proof, VerificationKey, groth16_verify}

pub type Datum {
  credential_hash: Scalar,   // Poseidon(pwd) as a BLS12-381 scalar
  vk: VerificationKey,       // compressed Groth16 verification key
}

pub type Redeemer {
  proof: Proof,              // Groth16 proof from snarkjs
}

validator password_lock {
  spend(
    datum: Option<Datum>,
    redeemer: Redeemer,
    _self: OutputReference,
    _ctx: Transaction,
  ) {
    expect Some(d) = datum
    groth16_verify(d.vk, redeemer.proof, [d.credential_hash])
  }

  else(_) { False }
}
\end{verbatim}

The datum stores two things: the Poseidon hash of the password (the lock) and the verification key (the lock mechanism). Both are public. The redeemer provides the proof (the key). The key never reveals the password.

\subsection*{Compiling}

From the \texttt{zk-password/} directory:

\begin{verbatim}
aiken check
\end{verbatim}

You should see:

\begin{verbatim}
Compiling chapter6/zk-password 0.0.0 (...)
Compiling aiken-lang/stdlib v2.1.0 (...)
\end{verbatim}

with zero errors.

% ─────────────────────────────────────────────────────────────────────────────
\section{Locking and Unlocking Funds} \label{sec:ch6-flow}

\subsection*{Locking}

To lock funds in the contract:

\begin{enumerate}
  \item Choose a password and compute \texttt{hash = Poseidon(pwd)} using \texttt{circomlibjs} in JavaScript.
  \item Run the trusted setup for \texttt{circuit.circom} to obtain the compressed verification key.
  \item Build a transaction that sends ADA to the script address derived from \texttt{password\_lock}, with a datum encoding \texttt{(hash, vk)}.
\end{enumerate}

The UTxO is now locked. Anyone can see the hash and the verification key on-chain, but without the password they cannot construct a valid proof.

\subsection*{Unlocking}

To unlock the funds:

\begin{enumerate}
  \item Generate \texttt{proof.json} with snarkjs using the password as the private input:
        \begin{verbatim}
snarkjs groth16 fullprove input.json circuit.wasm zkey_final.zkey \
    proof.json public.json
        \end{verbatim}
  \item Compress the proof points ($\pi_A$, $\pi_B$, $\pi_C$).
  \item Build a transaction spending the UTxO, with the redeemer encoding the compressed proof.
  \item Submit. The Aiken validator calls \texttt{groth16\_verify} and --- if the proof is valid --- approves the spend.
\end{enumerate}

\begin{highlight}
The password is used only \emph{locally} to generate the proof. It never appears in the transaction, the redeemer, or anywhere on-chain. An observer who inspects the blockchain sees the proof and the hash --- neither of which reveals the password.
\end{highlight}

% ─────────────────────────────────────────────────────────────────────────────
\section{How This Differs from Janus Wallet} \label{sec:ch6-vs-janus}

Janus Wallet is a full ZK wallet with staking, delegation, governance, and a server-side API. This chapter's contract is a stripped-down version that illustrates the same core idea without the surrounding infrastructure.

\begin{center}
\begin{tabular}{lll}
  \hline
  \textbf{Feature}          & \textbf{Janus Wallet}         & \textbf{zk-password} \\
  \hline
  Public inputs             & userId, challenge, hash, flags & hash only \\
  Replay protection         & On-chain challenge nonce       & None (educational) \\
  Server                    & Node.js + PostgreSQL           & Off-chain only \\
  Circuit language          & Circom + Poseidon              & Circom + Poseidon \\
  On-chain verifier         & Aiken + BLS12-381              & Aiken + BLS12-381 \\
  Staking / governance      & Yes                            & No \\
  \hline
\end{tabular}
\end{center}

\begin{remark}
The \texttt{zk-password} contract does not include a replay protection challenge. In production, any valid proof would allow spending, even if the same proof is reused. To prevent this, add a nonce or challenge to the circuit's public inputs, as Janus Wallet does with its \texttt{challenge} signal.
\end{remark}

This is the complete minimal ZK contract. In the next chapters we will add replay protection, multi-input circuits, and a proper off-chain proving workflow.
